National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-17 Manuscript.tex
revisions to manuscript, worked out practical effect sizes

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs},
  pdfauthor={Ben Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph Bulbulia5},
  pdflang={en-EN},
  pdfkeywords={Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Beliefs and Wellbeing}
\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 4778}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\usepackage{tcolorbox}
\usepackage{paralist}
\usepackage{multicol}
\usepackage{graphics}
\usepackage{rotating}
\usepackage{longtable}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi

\title{National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs}
\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
\date{}

@@

\abstract{
Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and reliance on cross-sectional samples cast doubts. Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective well-being.
}

@@

The relationship between spirituality and psychological well-being is a matter of enduring fascination (Ellsworth \& Ellsworth, 2010). Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005). However, critics raise three challenges.

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002). As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this features as a \enquote{tautology} in spirituality/mental health research (Koenig, 2008). Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employe tautological scales (Garssen et al., 2016). The first challenge, then, is to define measures of spirituality that are not well-being measures dressed in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001). However previous research does not converge on a unique set of meanings. Most studies of spirituality and religion have been conducted among (cross-sectional) North American samples among people who are affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012). However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being (King et al., 2013). Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangle the concept of spirituality from concepts of religion.

A third challenge is to investigate the relationship between spirituality and subjective well-being over the life-span in nationally diverse samples. To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out (Garssen \& Visser, 2016). However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012). To address the key questions in spirituality psychological health research, then, requires longitudinal investigations.

@@

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of N = 21,705 (N = 20,979 with complete responses).

@@

\textbf{Life Satisfaction} Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985). Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach's alpha for this 2-item scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

\textbf{Personal well-being} Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017). Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

\textbf{\emph{Belief in a Spirit or a God}} We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force. These believe were developed for the NZAVS from the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

@@

\textbf{Age} Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were \enquote{Male} was coded as \enquote{1} and \enquote{Female} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

\textbf{Education} Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree). Education was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

\textbf{Deprivation} We measured the socio-economic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014). The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80--90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

\textbf{Employed} Employment status was assessed by asking participants if they were currently working, \enquote{yes} was coded as \enquote{1} and \enquote{no} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

\textbf{Partner} Participants were asked if they were in a relationship, \enquote{yes} was coded as \enquote{1} and no was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

\textbf{Ethnicity} Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

\textbf{Urban} People were coded as either residing in an urban \enquote{1} or rural \enquote{0} area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.

\textbf{Political Orientation} To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative). Higher values indicate more conservative political beliefs. Political Orientation was standardized and centered at its mean. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

@@

\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Ids}
\end{aligned}
\]

@@

\hypertarget{personal-well-being}{%
\subsection{Personal Well-Being}\label{personal-well-being}}

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with Id as random effects. The model included Id as random effects. The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001) is at 6.63 (SE = 0.03, 95\% CI {[}6.57, 6.70{]}, p \textless{} .001). Within this model:

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs},
  pdfauthor={Ben Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph Bulbulia5},
  pdflang={en-EN},
  pdfkeywords={Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5015}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\usepackage{tcolorbox}
\usepackage{paralist}
\usepackage{multicol}
\usepackage{graphics}
\usepackage{rotating}
\usepackage{longtable}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi

\title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
\date{}

@@

\abstract{
Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
}

@@

The relationship between spirituality and psychological well-being is a matter of enduring fascination (Ellsworth \& Ellsworth, 2010). Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005). However, critics have raised three credible challenges.

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002). As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this approach as \enquote{tautological} (Koenig, 2008). Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employe tautological scales (Garssen et al., 2016). The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001). Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings. Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012). As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear. Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013). Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of contructs in populations where these dimensions overlap, partially overlap, and diverge.

A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans. Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assing how their lives turn out (Garssen \& Visser, 2016). Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012). However, To address the key questions in spirituality psychological health research requires longitudinal investigations. Unfortunately the relevant datasets are rare.

@@

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses).

@@

\hypertarget{life-satisfaction}{%
\subsubsection{Life Satisfaction}\label{life-satisfaction}}

Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985). Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach's alpha for this 2-item scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

\hypertarget{personal-well-being}{%
\subsubsection{Personal well-being}\label{personal-well-being}}

Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017). Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

\hypertarget{belief-in-a-spirit-or-a-god}{%
\subsubsection{Belief in a Spirit or a God}\label{belief-in-a-spirit-or-a-god}}

We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force. These believe were developed for the NZAVS from the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

@@

\hypertarget{age}{%
\subsubsection{Age}\label{age}}

Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were \enquote{Male} was coded as \enquote{1} and \enquote{Female} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

\hypertarget{education}{%
\subsubsection{Education}\label{education}}

Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree). Education was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

\hypertarget{deprivation}{%
\subsubsection{Deprivation}\label{deprivation}}

We measured the socio-economic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014). The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80--90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

\hypertarget{employed}{%
\subsubsection{Employed}\label{employed}}

Employment status was assessed by asking participants if they were currently working, \enquote{yes} was coded as \enquote{1} and \enquote{no} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

\hypertarget{partner}{%
\subsubsection{Partner}\label{partner}}

Participants were asked if they were in a relationship, \enquote{yes} was coded as \enquote{1} and no was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.

\hypertarget{ethnicity}{%
\subsubsection{Ethnicity}\label{ethnicity}}

Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

\hypertarget{urban}{%
\subsubsection{Urban}\label{urban}}

People were coded as either residing in an urban \enquote{1} or rural \enquote{0} area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.

\hypertarget{political-orientation}{%
\subsubsection{Political Orientation}\label{political-orientation}}

To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative). Higher values indicate more conservative political beliefs. Political Orientation was standardized and centered at its mean. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

@@

\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals}
\end{aligned}
\]

@@

\hypertarget{personal-well-being-1}{%
\subsection{Personal Well-Being}\label{personal-well-being-1}}

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, withId as random effects. The model included Id as random effects. The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001) is at 6.63 (SE = 0.03, 95\% CI {[}6.57, 6.70{]}, p \textless{} .001). Within this model:
